---
title: "Insurgent (Pairwise) Networks"
date: "Aug 22, 2022"
categories: [network analysis, R, combat]
description: "Conflict makes for strange bedfellows"
format:
  html:
    # embed-resources: true
    code-fold: true
    code-tools:
      source: false
      toggle: true
      caption: "Show the code"
    code-summary: "Show the code"
    # theme: flatly
    # theme: slate
    # keep-md: true
editor: visual
toc: true
number-sections: false
execute: 
  warning: false
  message: false
  error: false
  echo: true
---

During the height of the Syrian civil war, many of us watching from afar were puzzled when *"moderate"* insurgent groups aligned themselves with hardliners or jihadists. The conflict featured hundreds---perhaps thousands---of armed groups forming various kinds of alliances, only to break apart and later reconcile.

A lot is made of the role of ideology when armchair commanders (like myself) comment on these unfolding disasters. But alliances can also be pragmatic choices. By aggregating combat power, armed groups might increase their chances of survival---even if only to turn on each other at a later stage.

Recently, I came across a compelling argument in *Alliance Formation in Civil Wars by Fotini Christia.*[^1] In this book, Christia states: *"It would be natural to suppose that warring groups form alliances based on shared identity considerations---such as Christian groups allying with Christian groups or Muslim groups with their fellow co-religionists---but this is not what we see."*

[^1]: Fotini Christia, "[Alliance Formation in Civil Wars](https://www.amazon.com/Alliance-Formation-Professor-Fotini-Christia/dp/1107683483/ref=sr_1_1?keywords=alliance+formation+in+civil+wars&qid=1692661337&s=books&sprefix=alliance+formation%2Cstripbooks%2C100&sr=1-1)".

With that motivation in mind, I decided to explore what we could learn from Syrian armed group coalitions. When these groups engaged in battle, they frequently partnered with multiple other factions. This raised a key question for me: **Would groups tolerate fighting alongside ideological enemies?**

So, I focused on **pairwise relationships**, even within multi-group alliances.

```{r}
#| label: libraries

library(tidyverse)
library(ggraph)
library(tidygraph)
library(rsvg)
library(cowplot)
library(ggtext)
library(igraph)
library(devtools)
```

## Data

::: callout-important
## A love affair with ACLED data!

I came across the [Armed Conflict Location & Event Data Project (ACLED)](https://www.acleddata.com) a few years ago---and what an amazing resource.

ACLED is a disaggregated data collection, analysis, and crisis mapping project. The ACLED team collects both real-time and historical data on political violence and protest events across nearly 100 countries.

In short, it's an outstanding tool for understanding the dynamics of conflict. .[^2]
:::

[^2]: I accessed the [data from The Armed Conflict Location & Event Data Project (ACLED)](https://www.acleddata.com) in the spring of 2021.

ACLED data is robust, with many columns and coverage spanning multiple years. In this dataset, the `actor1` column (along with `assoc_actor_1`) represents entities that are acting together---typically in some form of violence---against `actor2`.

For this analysis, I'm only concerned with `actor1` and `assoc_actor_1`, since I'm interested in **collaborations**, not opponents. These columns capture groups that are working together during a reported event.

Let's take a quick peek at the data:

```{r}

read_csv("df_acled_syr_2017_2021.csv") |> 
  select(year, event_date, event_type, actor1, assoc_actor_1) |> 
  filter(!is.na(assoc_actor_1)) |>
  slice_head(n = 5) |> 
  knitr::kable()
```

I need select the columns of interest and then filter down to battle related events (filtering out `Protests`, `Riots`, and `Violence against civilians`)

```{r}

df_acled_syr <- 
  read_csv("df_acled_syr_2017_2021.csv") |> 
  filter(year %in% 2017:2019,
         event_type %in% c("Explosions/Remote violence",
                           "Battles",
                           "Strategic developments")) |> 
  select(actor1, assoc_actor_1) |> 
  mutate(actor1 = str_squish(actor1),
         assoc_actor_1 = str_squish(assoc_actor_1)) 
```

Here's the difficult part. The reduced ACLED data looks akin to this:

```{r}

(test <- tibble::tibble(
  actor1 = c("A", "A", "B"),
  assoc_actor_1 = c("B", "B; C", "C; D; E"))) |> 
  knitr::kable()
```

For what I'm interested in, I need to get the pairwise combinations. So, I need to work out and test the code to accomplish this. A huge thanks to [Dusty Turner](https://dustysturner.com/).[^3] In fact, this chunk is largely thanks to his generosity.

[^3]: Major [Dusty Turner](https://dustysturner.com/), U.S. Army is a monster coder and overall big brain. I've leaned on him for a number of problems. He consistently digs me out of a hole -- all while cracking a joke.

```{r}

#| # Separate assoc actors to get all pairwise partnerships'
df <- 
  test |> 
  # use only actors with multiple associated actors
  filter(str_detect(assoc_actor_1, ";")) |>
  separate_rows(assoc_actor_1, sep = ";") |>
  mutate(assoc_actor_1 = str_squish(assoc_actor_1)) |> 
  pivot_longer(actor1:assoc_actor_1) |> 
  select(value) |> 
  distinct(value) |> 
  mutate(value2 = value) |> 
  expand(value, value2) |> 
  filter(value !=value2) |> 
  mutate(helper = str_c(value,value2)) |> 
  rowwise() |> 
  mutate(helper = str_c(str_sort(unlist(str_split(helper, ""))),collapse = "")) |> 
  distinct(helper,.keep_all = T) |> 
  select(-helper) |> 
  rename(actor1 = value, assoc_actor_1 = value2)

# bind back with actors that don't have multiple assoc actors
test |> 
  filter(!str_detect(assoc_actor_1, ";")) |>
  bind_rows(df) |> 
  knitr::kable()

```

That looks good! Let's apply it to the ACLED data.

```{r}

# Separate assoc actors to get all pairwise partnerships'
df_separate <- 
  df_acled_syr |> 
  # use only actors with multiple associated actors
  filter(str_detect(assoc_actor_1, ";")) |>
  separate_rows(assoc_actor_1, sep = ";") |>
  mutate(assoc_actor_1 = str_squish(assoc_actor_1)) |> 
  pivot_longer(actor1:assoc_actor_1) |> 
  select(value) |> 
  distinct(value) |> 
  mutate(value2 = value) |> 
  expand(value, value2) |> 
  filter(value !=value2) |> 
  mutate(helper = str_c(value,value2)) |> 
  rowwise() |> 
  mutate(helper = str_c(str_sort(unlist(str_split(helper, ""))),collapse = "")) |> 
  distinct(helper,.keep_all = T) |> 
  select(-helper) |> 
  rename(actor1 = value, assoc_actor_1 = value2)

# bind back with actors that don't have multiple assoc actors
reshaped_df <- df_acled_syr |> 
  filter(!str_detect(assoc_actor_1, ";")) |>
  bind_rows(df_separate)
```

Ok, I have to confess. This next part stumped me and took forever. First, a huge shout out to Wikipedia contributors. There are soooo many actors in the dataset. I spend much more time than intended trying to consolidate them, constantly referencing Wikipedia to figure out what's going on.

Second, I needed to clean and consolidate the actors, and then remove the non-opposition groups as well as the non-armed-opposition groups. I tried several ways to make happen much more parsimoniously with `sapply` and `purrr::map` and a two-column lookup table, but it beat me. If you have suggestions, I'm eager to hear them.

```{r}

# create a lookup table for groups to remove
remove_groups <-
  c("Military Forces", "Police Forces", 
    "Operations Room", "Opposition Rebels", 
    "Alliance/Named Operation", "Tribal", 
    "Military Council", "Communal Militias", 
    "Civilians", "SDF", "Islamic State")

df_acled_actors <- 
  reshaped_df |>
  mutate(
    across(
      actor1:assoc_actor_1,
      ~ case_when(
        str_detect(.x, "Military Forces|Government|Allied Syrian") ~ "Military Forces",
        str_detect(.x, "Unidentified Armed|Opposition Rebels|Islamist|Sunni Muslim|JSH") ~ "Opposition Rebels",
        str_detect(.x, "Police Forces") ~ "Police Forces",
        str_detect(.x, "HXP|QSD|YPG|YPJ|Liberation Army of Afrin|Menbij Internal|Asayish|Syriac") ~ "SDF",
        str_detect(.x, "HTS|JFS|Jabhat Fateh al Sham") ~ "Hayat Tahrir al Sham",
        str_detect(.x, "Hamza Division|Hamza Brigade") ~ "Hamza Division", 
        str_detect(.x, "AAS:") ~ "Ahrar al Sham", 
        str_detect(.x, "Al Sham Corps|Al Sham Division") ~ "Faylaq al Sham",
        str_detect(.x, "HNDZ") ~ "Nour al Din al Zinki",
        str_detect(.x, "Sharqiya Army") ~ "Jaysh Sharqiya",
        str_detect(.x, "Liwa al Aqsa") ~ "Jund al Aqsa",
        str_detect(.x, "FaR:") ~ "Faylaq al Rahman",
        str_detect(.x, "JaS:") ~ "Levant Front",
        str_detect(.x, "Sultan Suleiman Shah") ~ "Sultan Suleiman Shah",
        str_detect(.x, "Operations Room") ~ "Operations Room",
        str_detect(.x, "Operation Room") ~ "Operations Room",
        str_detect(.x, "Wa Harredh al Moa'mineen") ~ "Operations Room",
        str_detect(.x, "JWS:|JTW:|JTS:|Euphrates Shield|Peace Spring") ~ "Alliance/Named Operation",
        str_detect(.x, "Islamic State") ~ "Islamic State",
        str_detect(.x, "Tribal") ~ "Tribal",
        str_detect(.x, "Military Council") ~ "Military Council",
        str_detect(.x, "TIP:") ~ "Turkistan Islamic Party",
        str_detect(.x, "Communal") ~ "Communal Militias",
        str_detect(.x, "Kurdish Ethnic") ~ "Kurdish Ethnic Militia",
        str_detect(.x, "JaT:") ~ "Army of the Revolutionaries",
        str_detect(.x, "Sultan Murad") ~ "Sultan Murad",
        str_detect(.x, "Ansar al Din") ~ "Ansar al Din",
        str_detect(.x, "JOS:") ~ "Lions of the East",
        str_detect(.x, "LAS") ~ "Northern Storm Brigade",
        str_detect(.x, "Civilians|Protesters|Rioters|Aid Workers|Women|Farmers|Refugees|Prisoners|Journalists|") ~ "Civilians",
        TRUE ~ .x
      ))) |>
  
  # remove civilans, state forces or unidentified groups
  filter(!actor1 %in% remove_groups,
         !assoc_actor_1 %in% remove_groups,
         actor1 != assoc_actor_1) |> 
  
  # Get the top/most groups by number of operations
  mutate(actor1 = fct_lump(actor1, 10),
         assoc_actor_1 = fct_lump(assoc_actor_1, 15)) |> 
  filter(actor1 != "Other",
         assoc_actor_1 != "Other")
```

Let's check out what the numbers look like for the top 10 most operationally-active groups. How many joint (aka partnered) operations does each group conduct - measured as a percentage of overall joint operations?

```{r }
library(DT)
datatable(df_acled_actors %>% 
  pivot_longer(cols = actor1:assoc_actor_1,
               names_to = "cols",
               values_to = "Name") %>%
  group_by(Name) %>% 
  summarize(`Partnered Ops` = n()) %>% 
  mutate(Percent = round(`Partnered Ops`/sum(`Partnered Ops`),2),
         Percent = scales::percent(Percent)) %>% 
  ungroup() %>% 
  arrange(desc(`Partnered Ops`))) 

```

Clearly, we expect Hayat Tahrir al-Sham to be a prominent player.

Now to prep the data for graphing...

```{r}

# create nodes with count of times each actor appears in dataset
df_nodes <- 
  df_acled_actors |> 
  pivot_longer(cols = actor1:assoc_actor_1,
               names_to = "cols",
               values_to = "Name") |> 
  group_by(Name) |> 
  summarize(count = n()) |> 
  ungroup() |> 
  #mutate(countG = cut(count, breaks = c(-Inf, 21, 24, 43, Inf))) |> 
  mutate(countG = case_when(
    count < 21 ~ "Least",
    count < 25 ~ "Less",
    count < 44 ~ "More",
    TRUE ~ "Most"
  ))

# from original df, weight is the count of how often the two actors work together
# igraph looks for "from", "to", and "weight"
df_edges <- 
  df_acled_actors |>
  count(actor1,assoc_actor_1) |> 
  rename(weight = n,
         from = actor1,
         to = assoc_actor_1)

graph <- 
  graph_from_data_frame(
  df_edges, 
  vertices = df_nodes)

```

And at last, let's generate a pairwise network plot. Let's see who partnered with whom.

```{r}
#| fig-height: 6
#| fig-width: 8

graphed <-
  graph |> 
  ggraph(layout = 'linear', 
         circular = TRUE) +
  ggraph::geom_edge_arc(
    aes(alpha = weight),
    width = 1,
    show.legend = FALSE,
    color = "grey20"
  ) +
  ggraph::geom_node_point(aes(color = countG)) +
  ggraph::geom_node_label(
    aes(label = name, 
        color = countG),
    alpha = .75,
    label.size = NA,
    fill = "#F3F3F3", 
    size = 3,
    repel = F,
    fontface = "bold"
  ) +
  labs(
    title = "Syrian Opposition: Never go it alone!",
    subtitle = "<b>Bolder lines</b> indicate more <i>joint</i> combat operations with that partner.
           <br>Colors indicate frequency of <i>total</i> combat operations:
           <br><b style='color:black'> Most Frequent</b>,
           <b style='color:#450D54'> More Frequent</b>,
           <b style='color:#557C9B'> Less Frequent</b>, and 
           <b style='color:#94B147'>Least Frequent</b>",
    caption = "Data:  <b>'ACLED Event Data 2017-2020'</b> (acleddata.com)<br> Visualisation by <b>Michael Davies</b>"
  ) +
  theme_void() +
  theme(
    # margins: top, right, bottom, and left
    plot.margin = margin(0.7, 0.7, 0.7, 0.7, "cm"),
    plot.title = element_text(size = 20, color = "grey30", face = "bold"),
    plot.title.position = 'plot',
    plot.subtitle = element_markdown(lineheight = 1.2),
    plot.caption = element_markdown(size = 8),
    plot.caption.position = 'plot',
    plot.background = element_rect(color = NA, fill = NA),
    legend.position = "none",
  ) +
  scale_color_manual(values = c("#94B147", "#557C9B", "#450D54", "black")) +
  coord_fixed(ratio = 0.6, clip = "off")


library(magick)
img <- 
  image_read("jihadi.PNG") |>
  image_resize("570x380") |>
  image_transparent("grey", fuzz = 35)

ggdraw() + 
  draw_plot(
    ggplot() + 
      theme_void() + 
      theme(
        plot.background = element_rect(color = NA, fill = "#F3F3F3") 
      )) +
  draw_image(interpolate = F, 
             img, 
             scale = .35, 
             x = 0, 
             y = 0) +
  draw_plot(graphed) 
```

Ahhh the tangled web they weave. I know I've been mixing metaphors.

I supposed I should have used the colors to signify ideology so that we can see if ideology and parnterships "travel well together." However, I became interested in how the frequency of operations and partnership might reflect battlefield success. For instance, Hayat Tahrir al-Sham, by all accounts, rose to be the most dominant opposition group. This is reflected in the fact that they conducted the most operations overall. Notably, however, they seldom engaged in an operation on their own. They engaged in the most frequent "joint operations" - despite being consider hardliners, jihadists, and ISIS offshoots.[^4]

[^4]: The Armed Conflict Location & Event Data Project (ACLED), [Actor Profile: Hayat Tahrir al-Sham (HTS) 26 July 2023](https://acleddata.com/2023/07/26/actor-profile-hayat-tahrir-al-sham-hts).

Maybe the key to insurgent survival is - never go it alone.
